Automated Lensing Learner - I: An automated Strong Lensing Identification Pipeline

### 2017-03-23 ms.tex
minor changes to matt's modifications

% Gravitational lensing as a geometric test of cosmology.  Also
% probes underlying matter distribution in galaxy clusters, sensitive
% to things like recent accretion and environment.
Gravitational lensing occurs when intermediate fluctuations in the
matter density field deflect light from background sources
\citep[see][for a review]{kneibandnatarajan_11}.  Strong gravitational
lensing can manifest as visible giant arcs magnifying high redshift
galaxies \citep{lyndsandpetrosian_86,gladders_etal03}, multiply imaged
quasars \citep{walsh_etal79}, multiply imaged galaxies
\citep{sharon_etal05}, and arclets \citep{bezecourt_etal98}.  Lensing
signatures probe the underlying dark matter distribution of the lens
\citep{warrenanddye_03}, and high redshift galaxy formation
\citep{allam_etal07}. They also provide a geometric test of cosmology via
comparison of predicted arc abundances with observed abundances
\citep{kochanek_96,chae_03,linder_04}, and via time-delay between signals
from multiply imaged quasars \citep{xli_etal12,suyu_etal14,suyu_etal16}.

@@

% Classical machine learning methods
Pattern recognition methods have transitioned to using ``machine
learning'' algorithms in place of manual cutoffs.  With
machine learning, we can train a model to separate a dataset into
a known set of classes, such as ``lensing systems'' and
``non-lensing systems''. However, many classic machine learning
algorithms do not work well in image space, i.e. directly using the
raw pixel values of the image, but first require a process of
\em{feature engineering}, in which domain expertise is used to
extract variables which are more directly related to the classification
task at hand.  An example would be the application of an edge detector.
The optimal weights and cutoffs for these derived variables that are
used to determine the class label of an image are found automatically
by the algorithm.

@@

% Convolutional and residual
The most advanced use of neural networks for strong lensing
classification operate directly in image space.
\citep{petrillo_etal17} used mock Kilo Degree Survey data to train
convolutional neural networks, with a training set size of six million
images. \citet{lanusse_etal17} used state-of-the-art deep residual
neural networks
to also work directly in the image space with minimal image
pre-processing.  A major strength of the \citet{lanusse_etal17}
implementation is that in comparison to deep convolutional neural
networks, deep residual neural networks have been found to
be easier to train and perform better.

%  Relevance of our work
While there has been a recent surge in the use of deep neural
networks applied to image classification problems in astronomy,
it is not always easy to scale these techniques to large data
sets, nor are the necessary computational resources and hardware,
such as GPUs, easily accessible to the
entire scientific community.  We present a supervised classification
pipeline built with open source tools, tested on mock HST and LSST
data, which will also be made publicly available.  Our goal is to
create a centralized open source pipeline that is general enough for
the user to select or add appropriate image processing techniques to
augment the features of the lensed image, train and test data with
different machine learning techniques, and to select the sequence of
methods that provides optimal model performance.  Results are
reproducible on personal computers, as both the pipeline and the
data will be publicly distributed.

@@

The first step of our pipeline consists of a feature extraction stage,
where our feature vector is a {\em histogram of oriented gradients}
(HOG) \citep{dalalandtriggs_05} that quantifies edges in the image.  We
describe the method and parameter search in Section~\ref{sec:hog}.  We
then use {\em Logistic Regression} (LR), a machine learning algorithm
described in Section~\ref{sec:LR}, to train a classifier model on a
subset of our images.  LR requires a parameter search over the
regression coefficient, $C_{LogReg}$, which we explore in
Section~\ref{sec:regularization}.  We briefly comment that our initial
tests with a {\em Support Vector Machine} (SVM) using radial basis
fucntions as an alternative
machine learning algorithm yielded negligible performance improvement,
and significantly increased computation cost.  This indicated that the
features of lens and non-lens images are relatively well separated by
hyperplanes in feature space.  For these reasons, we do not include
SVM in our final analysis and comparisons and continue all discussions
with a linear classifier.

>>>>

% Gravitational lensing as a geometric test of cosmology.  Also
% probes underlying matter distribution in galaxy clusters, sensitive
% to things like recent accretion and environment.
Gravitational lensing occurs when intermediate fluctuations in the
matter density field deflect light from background sources
\citep[see][for a review]{kneibandnatarajan_11}.  Strong gravitational
lensing can manifest as visible giant arcs magnifying high redshift
galaxies \citep{lyndsandpetrosian_86,gladders_etal03}, multiply imaged
quasars \citep{walsh_etal79}, multiply imaged galaxies
\citep{sharon_etal05}, and arclets \citep{bezecourt_etal98}.  Lensing
signatures probe the underlying dark matter distribution of the lens
\citep{warrenanddye_03}, and high redshift galaxy formation
\citep{allam_etal07}. Strong lenses also provide a geometric test of
cosmology via comparison of predicted arc abundances with observed
abundances \citep{kochanek_96,chae_03,linder_04}, and via time-delay
between signals from multiply imaged quasars
\citep{xli_etal12,suyu_etal14,suyu_etal16}.

@@

% Classical machine learning methods
Pattern recognition methods have transitioned to using ``machine
learning'' algorithms in place of manual cutoffs.  With machine
learning, we can train a model to separate a dataset into a known set
of classes, such as ``lensing systems'' and ``non-lensing
systems''. However, many classic machine learning algorithms do not
work well in image space, i.e. directly using the raw pixel values of
the image, but first require a process of \em{feature engineering}.
{\em Feature engineering} uses domain expertise 
to extract variables
that are more directly related to the classification task at hand.  An
example would be the application of an edge detector.  The optimal
weights and cutoffs for these derived variables that are used to
determine the class label of an image are found automatically by the
algorithm.

@@

% Convolutional and residual
The most advanced use of neural networks for strong lensing
classification operate directly in image space.
\citep{petrillo_etal17} used mock Kilo Degree Survey data to train
convolutional neural networks, with a training set size of six million
images. \citet{lanusse_etal17} used state-of-the-art deep residual
neural networks to also work directly in the image space with minimal
image pre-processing.  A major strength of the \citet{lanusse_etal17}
implementation is that in comparison to deep convolutional neural
networks, deep residual neural networks have been found to be easier
to train and perform better (Metcalf et al., in prep).

%  Relevance of our work
While there has been a recent surge in the use of deep neural networks
applied to image classification problems in astronomy, it is not
always easy to scale these techniques to large data sets, nor are the
necessary computational resources and hardware, such as graphical
processing units (GPUs), easily accessible to the entire scientific
community.  We present a supervised classification pipeline built with
open source tools.  We test the pipeline on mock HST and LSST data,
which will also be made publicly available.  Our goal is to create a
centralized open source pipeline that is general enough for the user
to select or add appropriate image processing techniques to augment
the features of the lensed image, train and test data with different
machine learning techniques, and to select the sequence of methods
that provides optimal model performance.  Results are reproducible on
personal computers, as both the pipeline and the data will be publicly
distributed.

@@

The first step of our pipeline consists of a feature extraction stage,
where our feature vector is a {\em histogram of oriented gradients}
(HOG) \citep{dalalandtriggs_05} that quantifies edges in the image.  We
describe the method and parameter search in Section~\ref{sec:hog}.  We
then use {\em Logistic Regression} (LR), a machine learning algorithm
described in Section~\ref{sec:LR}, to train a classifier model on a
subset of our images.  LR requires a parameter search over the
regression coefficient, $C_{LogReg}$, which we explore in
Section~\ref{sec:regularization}.  We briefly comment that our initial
tests with a {\em Support Vector Machine} (SVM) using radial basis
functions as an alternative
machine learning algorithm yielded negligible performance improvement,
and significantly increased computation cost.  This indicated that the
features of lens and non-lens images are relatively well separated by
hyperplanes in feature space.  For these reasons, we do not include
SVM in our final analysis and comparisons and continue all discussions
with a linear classifier.
